Evaluation of pilot jobs for Apache Spark\\ applications on HPC clusters

### 2019-05-20 paper.tex
final modifications

% Goal of the paper
In this paper, we study the use of pilot jobs to run scientific
applications with Apache Spark~\cite{zaharia2016apache} on shared HPC
clusters. The method currently recommended for this purpose involves batch
requesting all the
necessary resources and launching a Standalone Spark cluster once the
resources have been allocated. This is for instance the method used in
Compute Canada,
 our national computing infrastructure. With pilot
jobs, rather than requesting all the resources at once, a Spark cluster is
launched with a subset of the resources, and is expanded as more resources
get allocated. Consistently with previous examples of pilot job
deployments, we hypothesize that this strategy would reduce queuing times
by (1) fragmenting resource requirements, and (2) allowing shorter walltime
estimates. While there have been some efforts on implementing pilot jobs
for Apache Spark~\cite{jha-spark-pbs}, research is limited and
none of the them detail their quantitative effect.

Apache Spark is a popular Big Data framework, commonly used in both
industrial and academic settings. Although it is a Scala-based framework,
it also has APIs for Java, Python (PySpark) and R. Spark's Resilient
Distributed Dataset (RDD) abstraction enabled in-memory processing of
pipelines by co-locating tasks and data, which provided important
performance improvements compared to its predecessor Hadoop
MapReduce~\cite{dean2008mapreduce}. Through the use of RDDs, it also became possible
to execute iterative workflows -- something not easily doable in older
frameworks. Schedulers for Spark include its built-in standalone schedule,
Yet Another Resource Negotiator (YARN~\cite{apache13yet}), and Mesos~\cite{hindman2011mesos}. As a
result of the sustained growth in data volumes, Apache Spark and other
Big Data engines are increasingly used for scientific applications,
including in neuroimaging, our primary field of
interest~\cite{boubela2016big,mehta2017comparative,maybethesimulationone,freeman2014mapping}.

Big Data frameworks were designed with dedicated commodity infrastructure
in mind, and, with the exception of Dask~\cite{rocklin2015dask}, do not support batch
HPC schedulers such as PBS, SGE, Slurm and TORQUE~\cite{schedulers}, which
are commonly available to scientists. Therefore, to run Big Data
applications on HPC schedulers, it is also necessary to start an overlay
cluster that will schedule application tasks on resources provisioned
through batch schedulers. Our experiments quantify the effect of pilot jobs
when combined with such an overlay cluster. Our evaluation will be broken
down into two parts:
1) the added value of pilots, and 2), tolerance to failures.

@@

    A Spark cluster is made up of three components: the cluster manager, workers and
    driver. The cluster manager, also known as the Spark master in Standalone mode, is responsible for the
    resource provisioning within the cluster. Unlike the workers and driver, it is not
    application specific. The driver, which can start in the client where it is launched (i.e. client deploy mode)
    or directly on a worker node (i.e. cluster deploy mode) \TG{Deploy modes are explained in the next paragraph,
    I'd remove them from here}, must request resources from the master
    to launch its tasks. The master dispatches the requests to workers with the necessary resources available to start
    executors. The driver can then communicate with the executors to schedule and launch its tasks. Both the driver and
    the executors are application specific and connected directly to the master.

@@

    With HPC clusters, client mode is not necessarily possible on the login node as the
    driver client may require more resources than permitted. Moreover, due to cluster
    security policies, it may not be possible, or difficult to execute the driver from
    an external computer, such as a personal laptop. Therefore, for both batch and pilot
    scheduling, it is necessary to launch the driver from within the Slurm jobs, that is, in cluster mode.

    Spark also provides master fault tolerance through single-node recovery, where a shared
    filesystem is available. To enable this, a recovery folder must be set. Should the master
    fail, a new master will be able to takeover based on the information available in the recovery
    folder.

    In long running applications, it is necessary to be able to recovery the application execution from
    last successful state. Spark provides a checkpointing mechanism in order to ensure that the application
    is fault-tolerant to any kind of non application-related failure, such as a system failure.

@@

    There is not shortage of scripts available online for starting a Spark cluster on HPC, 
    see for instance \href{https://github.com/NIH-HPC/spark-slurm}{spark-slurm},
    \href{https://www.sherlock.stanford.edu/docs/software/using/spark}{scripts} provided
     by Stanford's Sherlock cluster, \href{https://sparkhpc.readthedocs.io}{sparkhpc},
     or \href{https://www.osc.edu/~troy/pbstools/man/pbs-spark-submit}{pbs-spark-submit}.
    However, all these scripts start up a cluster within a single batch call. At most, there are two calls, one to startup the cluster,
    and the second, to start the application execution. None of these use any kind of pilot-scheduling approach, nor do they discuss the
    effects of queuing time and how they might want to adjust it in the
    case where the driver is started separately from the batch job to start
    the cluster. Even in the case of RADICAL-Pilot, as per some available
    tutorials for using the pilot-job application with
    Spark (see
    \href{https://github.com/radical-cybertools/pilot-streaming/blob/master/examples/Pilot-Streaming-GettingStarted.ipynb}{here}
    and \href{https://github.com/radical-cybertools/MIDAS-tutorial/blob/master/pilot/Pilot-Spark.ipynb}{here}),
    it appears that the Spark cluster is started within a single pilot and
    the application is submitted, in a separate process, to that pilot. Due
    to the fact that existing solutions mainly rely on a single batch call
    to startup the Spark cluster and do not appear to be able to submit
    multiple smaller jobs to improve queuing times, we have decided to
    implement our own solution for the sake of our experiments.

@@

    The pilot submission template is similar to that of the batch template, however,
    each pilot will start its own Spark master and worker. However, there will only be one
    pilot which will start the driver process. The pilot selected to start the driver is the first one
    that attempts to do so by way of lockfile.
    The reason for 
    which each pilot starts its own master is to ensure the fault tolerance of the 
    masters. In this sort of configuration, should the active master be killed, one
    of the stand-by masters will takeover and the application will be able to 
    resume. Such a configuration is particularly favourable in pilot scheduling 
    scenarios as node failures may be more frequent due to walltime expiration.
    Additionally, the Spark deploy mode of the driver was selected to be cluster deploy mode.
    This would not only allow the driver to be executed directly on one of the workers,
    but also allow us to make the driver fault-tolerant through the
    \texttt{supervise} mode, which is only available in cluster deploy. As
    with the masters, it is particularly important to have a fault-tolerant
    driver in pilot-scheduling scenarios due to possible walltime
    expiration.
    Should pilots be idle for a certain duration, the
    pilots will terminate themselves such as to not hog resources.
    % \TG{It should be explained before that one of the interests
    % of pilots is to ``play'' with shorter walltimes}. \TG{It would also be
    % nice to have runs where pilots 
    % have shorter walltimes, to see if queuing times are shorter too.}
    \TG{You could add a figure showing the driver, workers, master and stand-by masters, to make the system 
    description clearer.}

@@

    Investigating the fault-tolerance provided by the overlay cluster is
    crucial in determining the suitability of pilot scheduling with overlay clusters \TG{You should water this down if 
    no fault tolerance experiment is presented}.
    Fault-tolerance of the driver is only possible in cluster deploy mode, however,
    when using Spark's Standalone scheduler, this mode is not available for Python 
    applications. It is for this reason that our synthetic application is written
    in Scala. Nevertheless, cluster deploy mode is possible for Python application
    using YARN or Mesos schedulers.

@@

       There were 15 repetitions in total for each configuration, and the order
       in which the configurations were launched was also randomized. This was to 
       account for any system variability that can occur, particularly in production
       HPC clusters.

       Different clusters may have different resource configurations, number of 
       users and scheduling policies. Therefore, we executed all 15 repetitions on
       both Cedar and B\'eluga to determine how much our results differ between
       two distinct clusters.

@@

As can be seen in both Figures~\ref{fig:makespansbeluga} and~\ref{fig:makespanscedar}, pilots 
generally did not bring any performance improvement to the execution of Spark workloads
on either system. In B\'eluga and Cedar, 16 pilots were generally slower than batch except for a 
few occurrencs. When 16 pilots were faster, it was generally either very slight, or, as expected,
when batch queuing times were very large. The maximum speedup achieved
was approximately 5.6~x for 8 and 16 pilots on B\'eluga and 2.6~x on Cedar. 
It is only in Configuration~3 (Figure~\ref{fig:beluga4})
that we spot any large speedups, however, it generally remains that
pilots are slightly slower, even in Configuration~3. The same can be said about Cedar, however,
it occured with Configuration~4 instead of 3. On average, pilots were up to 0.11~x slower on B\'eluga and
0.17~x slower on Cedar (see Table~\ref{table:speedup}), both of which occurred when using 16 pilots.
The only times when pilots were found to be, on average, faster, was in Configuration~3 on B\'eluga
and Configuration~4 on Cedar where the speedup was found to reach a max of 1.46 for 8 pilots on 
B\'eluga. The missing bars in Figures~\ref{fig:makespansbeluga} and~\ref{fig:makespanscedar} correspond to 
failed executions.

@@

In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
are used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
given an experiment configuration, where the line corresponds to the estimated makespan using 
our performance model. Makespan for the model was calculated as:
$$
M = \frac{125\times(d + 20)}{W}\times 10,
$$
where:
\begin{itemize}
    \item $125$ is the number of chunks BigBrain was split into
    \item $M$ is the makespan of the application, in seconds
    \item $W$ is the average number of Spark workers throughout the execution, computed as in Equation~\ref{eq:avgw}
    \item $d$ is the task delay, in seconds, associated with a given configuration
    \item $20$ is the average measured incrementation duration, in seconds, for a BigBrain chunk
    \item $10$ is the number of iterations
\end{itemize}
As can be seen in the Figure~\ref{fig:mwall}, both pilots and batch 
followed the general trend denoted by the model, which confirms that 
the system was behaving without major external sources of
overhead. However, pilots were consistently slower than batch for the same
number of average workers. This is particularly visible in Configuration 4
on B\'eluga (Fig. \ref{fig:mwbeluga}), and in Configuration 2, 3 and 4 on
Cedar (Fig. \ref{fig:mwcedar}). It means that some sort of overhead
impacted pilots but not batch. It can also be noticed that both batch and
pilots occasionally deviated from the model line, in particular
Configurations 2 and 3 in B\'eluga: this is due to the fact that our
application is not a divisible load, as the model assumes. Extending the
model beyond divisible loads is easy enough for batch and confirms that the
observed deviations from the model line come from this assumption (see gray
line in Fig.~\ref{fig:mwall} \TG{can you add the step line in gray? I
realize that this is not the best explanation, but I don't have the
expression for pilots. Also, I think this comment belongs here as it directly impacts
the reading of the results, but it might as well be moved to the discussion if you want to keep
this section short and sweet.}). The extended expression for pilots is more complex
though.

When
comparing the average number of worker difference
(Figures~\ref{fig:nworkersbeluga} and~\ref{fig:nworkerscedar}), it can be
seen that (1) repetitions where pilots were faster than batch correspond
to repetitions where pilots had more workers, and vice versa -- this
confirms that performance differences are mainly coming from queuing time
differences, and (2) pilot average number of workers start to exceed batch
as the application requirements increase, that is, for Configuration 3 and
4. However, employing 16 pilots
is generally always more costly than 8 pilots \TG{The last sentence looks off-topic to me but maye I didn't understand.}.

    \begin{figure*}
        \centering
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_1_beluga}
            \caption[]%
            {{\small Configuration 1}}
            \label{fig:nwbeluga1}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_2_beluga}
            \caption[]%
            {{\small Configuration 2}}
            \label{fig:nwbeluga2}
        \end{subfigure}
        \vskip\baselineskip
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_3_beluga}
            \caption[]%
            {{\small Configuration 3}}
            \label{fig:nwbeluga3}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_4_beluga}
            \caption[]%
            {{\small Configuration 4}}
            \label{fig:nwbeluga4}
        \end{subfigure}
        \caption[]
        {\small  The difference in average workers on B\'eluga between batch and pilots for all configurations and repetitions. Positive values mean
        that batch had more workers than pilots, that is, pilots did not improve queuing times.}
        \label{fig:nworkersbeluga}
    \end{figure*}
    \begin{figure*}
        \centering
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_1_cedar}
            \caption[]%
            {{\small Configuration 1}}
            \label{fig:nwcedar1}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_2_cedar}
            \caption[]%
            {{\small Configuration 2}}
            \label{fig:nwcedar2}
        \end{subfigure}
        \vskip\baselineskip
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_3_cedar}
            \caption[]%
            {{\small Configuration 3}}
            \label{fig:nwcedar3}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_4_cedar}
            \caption[]%
            {{\small Configuration 4}}
            \label{fig:nwcedar4}
        \end{subfigure}
        \caption[]
        {\small The difference in average workers on Cedar between batch and pilots for all configurations and repetitions. Positive values mean
        that batch had more workers than pilots, that is, pilots did not improve queuing times.}
        \label{fig:nworkerscedar}
    \end{figure*}

\begin{table}                                                                    
    \centering                                                                       
    \begin{tabular}{c|c|c|c|c}                                                             
        {} & \multicolumn{2}{c}{B\'eluga} & \multicolumn{2}{c}{Cedar}\\
    \rowcolor{headcolor}                                                             
    Configuration & 8 pilots & 16 pilots & 8 pilots & 16 pilots\\                               
    \hline                                                                           
    1 & 0.949 & 0.932 & 0.724 & 0.874\\                                               
    2 & 0.988 & 0.916 & 0.836 & 0.826\\                                               
    3 & 1.458 & 1.369 & 0.941 & 0.964\\
    4 & 0.970 & 0.891 & 1.049 & 1.068\\
    \end{tabular}                                                                    
    \setlength{\belowcaptionskip}{-10pt}                                             
    \caption{Speedup of pilots compared to batch}                                                    
    \label{table:speedup}                                                            
\end{table}

@@

%Technical overhead of using pilots + an overlay cluster over a cluster.
Pilots do not appear to provide any performance advantage over batch
scheduling to deploy Apache Spark overlay clusters on HPC.
This can be due to a few reasons. The queuing time for batch and pilots were similar, 
which can be an indication that we were just lacking the right level of priority to 
have been scheduled immediately. In the instances where batch was found to be significantly slower,
it might have been then case that we had obtained the necessary priority level but there
were not enough resources available to be scheduled.

@@

The motivation to use pilots, however, is dynamic scheduling on HPC clusters. While we
investigated the queuing time differences with respect to differences in available resources,
the walltimes were kept static. Had walltime been underestimated to the point where the Slurm jobs
would terminate prior to application completion, pilots might have been preferable to batch
as batch scheduling has no mechanism to restart. Even if such a mechanism was in place, the entire
cluster would be shutdown and need to restart, which has additional overheads. With pilots, as long as 
there was at least one pilot alive, it would be possible to maintain the cluster and add additional pilots.
Spark provides built-it fault-tolerance for not only the workers, but the master and driver as well. However,
it is also likely that with faster queuing times, all resources would be allocated at the same time. If walltime
were to be underestimated, in this case, checkpointing and restarting the entire application from the last
checkpoint would be the only option. This too has limitations as the last checkpoint may not be recent, requiring
necessitating significant recomputation. It
might even be desirable to have many small resource requests in such a scenario, as with an increase in pilots,
it becomes less likely for all of them to be scheduled at once, as can be seen with 16 pilots. This would allow
for more chances for the application to reach completion without failing entirely.

Many failures can be found, particularly with regards to the pilots. Many
of these have been a result of Spark failing to start properly. The driver
either would not return a submission ID after being started, or would
remain in \texttt{SUBMITTED} status until the walltime expiration, having
not processed any data. As these error become fewer with increased
resources, it is possible they are entirely related to lack of resources.
Furthermore, it may also explain why batch submission always passes. Not
only are all the workers that were requested available when the pilot
begins running, but batch also has one extra executor available thanks to the driver
process running directly on the process which had created it, rather than on a
worker process. Conversely, it is not only the pilot applications that
fail, the batch applications do experience some failures as well. It can be
seen in Figure~\ref{fig:makespansbeluga} that the batch application failed
twice: once in Configuration~2 and the other in Configuration~4. In both
cases, the reason for failure was the same: lack of available resources
despite all executors having been assigned. These failures might be due to
running on a shared HPC cluster. Either the resources were down, or other users might
have been using them, resulting in the resources being unavailable to our
application.

Running a Spark cluster atop of a Slurm allocation complexifies debugging, particularly if the application runs in cluster mode.
Spark worker logs, unless saved to network storage, are unavailable. Saving them to network storage also effectively slows down
the application as it is a slower storage than RAM and local disk. As a result, the Spark worker UI is also inaccessible. When running
in client mode, the driver output ends up it standard output/error. However, in cluster mode, the driver output is found stored within
the worker logs, which means the driver logs are also inaccessible without worker logs being written to network storage. Some HPC clusters
may have a file limit for user, meaning that the user must be sure that these logs coupled without other stored data does not exceed the file
limit.

\TG{You could end the discussion by a comment on the current literature on pilot jobs. How are your results
consistent or not with the literature?}

>>>>

% Goal of the paper
In this paper, we study the use of pilot jobs to run scientific
applications with Apache Spark~\cite{zaharia2016apache} on shared HPC
clusters. The method currently recommended for this purpose involves batch
requesting all the
necessary resources and launching a Standalone Spark cluster once the
resources have been allocated. This is, for instance, the method used on
Compute Canada,
 our national computing infrastructure. With pilot
jobs, rather than requesting all the resources at once, a Spark cluster is
launched with a subset of the resources, and is expanded as more resources
get allocated. Consistently with previous examples of pilot job
deployments, we hypothesize that this strategy would reduce queuing times
by (1) fragmenting resource requirements, and (2) allowing shorter walltime
estimates. While there have been some efforts on implementing pilot jobs
for Apache Spark~\cite{luckow2016hadoop}, research is limited and
none of the them detail their quantitative effect.

Apache Spark is a popular Big Data framework, commonly used in both
industrial and academic settings. Although it is a Scala-based framework,
it also has APIs for Java, Python (PySpark) and R. Spark's Resilient
Distributed Dataset (RDD) abstraction enabled in-memory processing of
pipelines by co-locating tasks and data, which provided important
performance improvements compared to its predecessor Hadoop
MapReduce~\cite{dean2008mapreduce}. Through the use of RDDs, it also became possible
to execute iterative workflows -- something not easily doable in older
frameworks. Schedulers for Spark include its built-in standalone schedule,
Yet Another Resource Negotiator (YARN~\cite{apache13yet}), and Mesos~\cite{hindman2011mesos}. As a
result of the sustained growth in data volumes, Apache Spark and other
Big Data engines are increasingly used for scientific applications,
including in neuroimaging, our primary field of
interest~\cite{boubela2016big,mehta2017comparative,freeman2014mapping}.

Big Data frameworks were designed with dedicated commodity infrastructure
in mind, and, with the exception of Dask~\cite{rocklin2015dask}, do not support batch
HPC schedulers such as the \href{http://www.univa.com/products/}{Oracle Grid Engine (OGE)},
\href{https://slurm.schedmd.com/}{Slurm} and \href{https://www.adaptivecomputing.com/products/torque/}{TORQUE}, which
are commonly available to scientists. Therefore, to run Big Data
applications on HPC schedulers, it is also necessary to start an overlay
cluster that will schedule application tasks on resources provisioned
through batch schedulers. Our experiments quantify the effect of pilot jobs
when combined with such an overlay cluster.

@@

    A Spark cluster is made up of three components: the cluster manager, workers and
    driver. The cluster manager, also known as the Spark master in Standalone mode, is responsible for the
    resource provisioning within the cluster.  As it is the resource manager for a given cluster,
    the cluster manager is not application specific. The driver requests resources from the master
    to launch its tasks. The master then dispatches the requests to workers with
    the necessary resources available to start
    executors. The driver may subsequently communicate with the executors to schedule and launch its tasks. Both the driver and
    the executors are application specific and connected directly to the master.

@@

    With HPC clusters, client mode is not necessarily possible on the login node as the
    driver client may require more resources than permitted. Moreover, due to cluster
    security policies, it may be difficult or impossible, to execute the driver from
    an external computer, such as a personal laptop. Therefore, for both batch and pilot
    scheduling, it is necessary to launch the driver from within the Slurm jobs. That is, in cluster mode.

    Spark also provides master fault tolerance through single-node recovery, wherever a shared
    filesystem is available. To enable this, a recovery folder must be set. Should the master
    fail, a new master will be able to takeover based on the information available in the recovery
    folder.

    In long running applications, it is necessary to be able to recover application execution from
    last successful state. Spark provides a checkpointing mechanism in order to ensure that the application
    is fault-tolerant to any kind of non application-related failure, such as a system failure.

@@

    There is no shortage of scripts available online for starting a Spark cluster on HPC, 
    see for instance \href{https://github.com/NIH-HPC/spark-slurm}{spark-slurm},
    \href{https://www.sherlock.stanford.edu/docs/software/using/spark}{scripts} provided
     by Stanford's Sherlock cluster, \href{https://sparkhpc.readthedocs.io}{sparkhpc},
     or \href{https://www.osc.edu/~troy/pbstools/man/pbs-spark-submit}{pbs-spark-submit}.
    However, all these scripts start up a cluster within a single batch call. At most, there are two calls, one for
    cluster startup,
    and the other, to launch the application. None of these use any kind of pilot-scheduling approach, nor do they discuss the
    effects of queuing time and how they might want to adjust it in the
    case where the driver is started separately from the batch job to start
    the cluster. Even in the case of RADICAL-Pilot, as per some available
    tutorials for using the pilot-job application with
    Spark (see
    \href{https://github.com/radical-cybertools/pilot-streaming/blob/master/examples/Pilot-Streaming-GettingStarted.ipynb}{here}
    and \href{https://github.com/radical-cybertools/MIDAS-tutorial/blob/master/pilot/Pilot-Spark.ipynb}{here}),
    it appears that the Spark cluster is started within a single pilot and
    the application is submitted, in a separate process, to that pilot. Due
    to the fact that existing solutions mainly rely on a single batch call
    to startup the Spark cluster and do not appear to be able to submit
    multiple smaller jobs to improve queuing times, we have decided to
    implement our own solution for the sake of our experiments.

@@

    The pilot submission template is similar to that of the batch template, although,
    each pilot will start its own Spark master and worker. However, there will only be one
    pilot which will start the driver process. The pilot selected to start the driver is the first one
    that attempts to do so by way of lockfile.
    The reason for 
    which each pilot starts its own master is to ensure the fault tolerance of the 
    masters. In this configuration, should the active master be killed, one
    of the stand-by masters can takeover and the application may be able to 
    resume if single-node recovery is set. Such a configuration is particularly favourable in pilot scheduling 
    scenarios as node failures may be more frequent due to walltime expiration.
    Additionally, the Spark deploy mode of the driver was selected to be cluster deploy mode.
    This would not only allow the driver to be executed directly on one of the workers,
    but also allow us to make the driver fault-tolerant through the
    \texttt{supervise} mode, which is only available in cluster deploy. As
    with the masters, it is particularly important to have a fault-tolerant
    driver in pilot-scheduling scenarios due to possible walltime
    expiration.
    Should pilots be idle for a certain duration, the
    pilots will terminate themselves such as to not hog resources.
    % \TG{It should be explained before that one of the interests
    % of pilots is to ``play'' with shorter walltimes}. \TG{It would also be
    % nice to have runs where pilots 
    % have shorter walltimes, to see if queuing times are shorter too.}

@@

    Spark cluster fault-tolerance is important in determining the suitability of
    pilot-jobs for Spark applications on HPC. Executing jobs on a shared cluster may
    result in a variety of failures. Pilot jobs may be more likely to fail as a result of
    underestimation of resources (e.g. walltimes) and therefore should be fault-tolerant
    to master, worker and driver failures. Although built-in fault-tolerance was not
    investigated in this paper, we want to ensure that the environment is set up in such
    a way that Spark's fault-tolerance configuration could easily be set should our 
    experiments return favourable results for pilot-jobs.
    Fault-tolerance of the driver is only possible in cluster deploy mode, however,
    when using Spark's Standalone scheduler, this mode is not available for Python 
    applications. It is for this reason that our synthetic application is written
    in Scala. Nevertheless, cluster deploy mode is possible for Python application
    using YARN or Mesos schedulers.

@@

       There were 10 repetitions in total for each configuration, and the order
       in which the configurations were launched was also randomized. This was to 
       account for any system variability that can occur, particularly in production
       HPC clusters.

       Different clusters may have different resource configurations, number of 
       users and scheduling policies. Therefore, we executed all 10 repetitions on
       both Cedar and B\'eluga to determine how much our results differ between
       two distinct clusters.

@@

As can be seen in both Figures~\ref{fig:makespansbeluga} and~\ref{fig:makespanscedar}, pilots 
generally did not bring any performance improvement to the execution of Spark workloads
on either system. In B\'eluga and Cedar, 16 pilots were generally slower than batch except for a 
few occurrencs. When 16 pilots were faster, it was generally either very slight, or, as expected,
when batch queuing times were very large. The maximum speedup achieved
was approximately 5.6~x for 8 and 16 pilots on B\'eluga and 2.6~x on Cedar. 
It is only in Configuration~3 (Figure~\ref{fig:beluga4}), on B\'eluga,
that we spot any large speedups, however, it generally remains that
pilots are slightly slower, even in Configuration~3. The same can be said about Cedar, however,
it occured with Configuration~4 instead of 3. On average, pilots were up to 0.11~x slower on B\'eluga and
0.17~x slower on Cedar (see Table~\ref{table:speedup}), both of which occurred when using 16 pilots.
The only times when pilots were found to be, on average, faster, was in Configuration~3 on B\'eluga
and Configuration~4 on Cedar where the speedup was found to reach a max of 1.46 for 8 pilots on 
B\'eluga. The missing bars in Figures~\ref{fig:makespansbeluga} and~\ref{fig:makespanscedar} correspond to 
failed executions.

@@

In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
were used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
given an experiment configuration, where the line corresponds to the estimated makespan using 
our performance model. Makespan for the model was calculated as:
$$
M = \frac{125\times(d + 20)}{W}\times 10,
$$
where:
\begin{itemize}
    \item $125$ is the number of chunks BigBrain was split into
    \item $M$ is the makespan of the application, in seconds
    \item $W$ is the average number of Spark workers throughout the execution, computed as in Equation~\ref{eq:avgw}
    \item $d$ is the task delay, in seconds, associated with a given configuration
    \item $20$ is the average measured incrementation duration, in seconds, for a BigBrain chunk
    \item $10$ is the number of iterations
\end{itemize}
As can be seen in the Figure~\ref{fig:mwall}, both pilots and batch 
followed the general trend denoted by the model, which confirms that 
the system was behaving without major external sources of
overhead. However, pilots were consistently slower than batch for the same
number of average workers. This is particularly visible in Configuration 4
on B\'eluga (Fig. \ref{fig:mwbeluga}), and in Configuration 2, 3 and 4 on
Cedar (Fig. \ref{fig:mwcedar}). It means that some sort of overhead
impacted pilots but not batch. It can also be noticed that both batch and
pilots occasionally deviated from the model line, in particular
Configurations 2 and 3 in B\'eluga: this is due to the fact that our
application is not a divisible load, as the model assumes. Extending the
model beyond divisible loads is easy enough for batch and confirms that the
observed deviations from the model line come from this assumption (see gray
line in Fig.~\ref{fig:mwall}. The extended expression for pilots is more complex
though.

When
comparing the average number of worker difference
(Figures~\ref{fig:nworkersbeluga} and~\ref{fig:nworkerscedar}), it can be
seen that (1) repetitions where pilots were faster than batch correspond
to repetitions where pilots had more workers, and vice versa -- this
confirms that performance differences are mainly coming from queuing time
differences, and (2) pilot average number of workers start to exceed batch
as the application requirements increase, that is, for Configuration 3 and
4.

\begin{figure*}

        \centering
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_1_beluga}
            \caption[]%
            {{\small Configuration 1}}
            \label{fig:nwbeluga1}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_2_beluga}
            \caption[]%
            {{\small Configuration 2}}
            \label{fig:nwbeluga2}
        \end{subfigure}
        \vskip\baselineskip
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_3_beluga}
            \caption[]%
            {{\small Configuration 3}}
            \label{fig:nwbeluga3}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_4_beluga}
            \caption[]%
            {{\small Configuration 4}}
            \label{fig:nwbeluga4}
        \end{subfigure}
        \caption[]
        {\small  The difference in average workers on B\'eluga between batch and pilots for all configurations and repetitions. Positive values mean
        that batch had more workers than pilots, that is, pilots did not improve queuing times.}
        \label{fig:nworkersbeluga}
    \end{figure*}
    \begin{figure*}
        \centering
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_1_cedar}
            \caption[]%
            {{\small Configuration 1}}
            \label{fig:nwcedar1}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_2_cedar}
            \caption[]%
            {{\small Configuration 2}}
            \label{fig:nwcedar2}
        \end{subfigure}
        \vskip\baselineskip
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_3_cedar}
            \caption[]%
            {{\small Configuration 3}}
            \label{fig:nwcedar3}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_4_cedar}
            \caption[]%
            {{\small Configuration 4}}
            \label{fig:nwcedar4}
        \end{subfigure}
        \caption[]
        {\small The difference in average workers on Cedar between batch and pilots for all configurations and repetitions. Positive values mean
        that batch had more workers than pilots, that is, pilots did not improve queuing times.}
        \label{fig:nworkerscedar}
    \end{figure*}

\begin{table}                                                                    
    \centering                                                                       
    \begin{tabular}{c|c|c|c|c}                                                             
        {} & \multicolumn{2}{c}{B\'eluga} & \multicolumn{2}{c}{Cedar}\\
    \rowcolor{headcolor}                                                             
    Configuration & 8 pilots & 16 pilots & 8 pilots & 16 pilots\\                               
    \hline                                                                           
    1 & 0.949 & 0.932 & 0.724 & 0.874\\                                               
    2 & 0.988 & 0.916 & 0.836 & 0.826\\                                               
    3 & 1.458 & 1.369 & 0.941 & 0.964\\
    4 & 0.970 & 0.891 & 1.049 & 1.068\\
    \end{tabular}                                                                    
    \setlength{\belowcaptionskip}{-10pt}                                             
    \caption{Average speedup of pilots for each configuration}                                                    
    \label{table:speedup}                                                            
\end{table}

@@

%Technical overhead of using pilots + an overlay cluster over a cluster.
Pilots do not appear to provide any performance advantage over batch
scheduling when deploying Apache Spark overlay clusters on HPC.
This can be due to a few reasons. The queuing time for batch and pilots were similar, 
which can be an indication that we were just lacking the right level of priority to 
have been scheduled immediately. In the instances where batch was found to be significantly slower,
it might have been case that we had obtained the necessary priority level but there
were not enough resources available to be scheduled.

@@

The motivation to use pilots, however, is dynamic scheduling on HPC clusters. While we
investigated the queuing time differences with respect to differences in available resources,
the walltimes were kept static. Had walltime been underestimated to the point where the Slurm jobs
would terminate prior to application completion, pilots might have been preferable to batch
as batch scheduling has no mechanism to restart. Even if such a mechanism was in place, the entire
cluster would be shutdown and need to restart, which has additional overheads. With pilots, as long as 
there was at least one pilot alive, it would be possible to maintain the cluster and add additional pilots.
Spark provides built-it fault-tolerance for not only the workers, but the master and driver as well. However,
it is also likely that with faster queuing times, all resources would be allocated at the same time. If walltime
were to be underestimated, in this case, checkpointing and restarting the entire application from the last
checkpoint would be the only option. This too has limitations as the last checkpoint may not be recent,
necessitating significant amounts of recomputation. It
might even be desirable to have many small resource requests in such a scenario, as it becomes less likely for all pilots to start at once as they become more numerous, as can be seen with 16 pilots when compared to 8. This would allow
for more chances for the application to reach completion without failing entirely.

Many failures can be found, particularly with regards to the pilots. Most
of these have been a result of Spark failing to start properly. The driver
either would not return a submission ID after being started, or would
remain in \texttt{SUBMITTED} status until the walltime expiration, having
not processed any data. As these errors become fewer with increased
resources, it is possible they are entirely related to lack of resources.
Furthermore, it may also explain why batch submission is more frequently successful. Not
only are all the workers that were requested available when the pilot
begins running, but batch also has one extra executor available thanks to the driver
process running directly on the process which had created it, rather than on a
worker process. Conversely, it is not only the pilot applications that
fail, the batch applications do experience some failures as well. It can be
seen in Figure~\ref{fig:makespansbeluga} that the batch application failed
twice: once in Configuration~2 and the other in Configuration~4. In both
cases, the reason for failure was the same: lack of available resources
despite all executors having been assigned. These failures might be due to
running on a shared HPC cluster. Either the resources were down, or other users might
have been using them, resulting in the resources being unavailable to our
application.

Running a Spark cluster atop of a Slurm allocation complexifies debugging, particularly if the application runs in cluster mode.
Spark worker logs, unless saved to network storage, become unavailable after program execution.
Although the worker UI is available during program execution, a user must manually create an SSH tunnel
to each worker node to gain access to them.
An option is to save the to a shared network storage for future access. This effectively slows down
the application as the network FS is typically a slower storage than local disk and RAM.

When running in client mode, the driver output ends up it standard output/error. 
However, in cluster mode, the driver output is found stored within
the worker logs, which means the driver logs are also inaccessible without 
worker logs being written to network storage. Some HPC clusters
may have a file limit for users, meaning that the user must be sure that these 
logs coupled without other stored data does not exceed the file
limit.
